---
title: "lab3"
output:
  pdf_document: default
  html_document: default
date: "2026-02-05"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Q1

## a
Lets choose a uniform distribution instead of a normal envelope for simplicity. The triangle density needs an envelope in [-1, 1], elsewhere it is 0. It has a maxima of 1 so we need to scale our uniform envelope by 2. Our waste will be 1/2, we enclose a function with area 1 with a function of area 2.
```{r 1_1_1, echo=FALSE, warning=FALSE}

tri_fun <- function(x) {
  
  
  
  
  if (x < -1) {
    return(0)
  } else if (x <= 0) {
    return(x+1)
  } else if (x <= 1) {
    return(1-x)
  } else {
    return(0)
  }
  
}

y <- sapply(seq(-2, 2, length.out = 1000), tri_fun)
plot(seq(-2, 2, length.out = 1000), y)

```



```{r 1_1_2, echo=FALSE, warning=FALSE}

set.seed(123)
n <- 10000
rejection_count <- 0
sample <- c()

while(length(sample) < n){
  U1 <- runif(1, -1, 1)
  U2 <- runif(1, 0, 1)
  
  
  
  if (tri_fun(U1) > U2) {
      
      sample <- c(sample, U1)

  } else {
    rejection_count <- rejection_count + 1
  }
}

waste <- rejection_count / (n + rejection_count)
print(waste)

hist(sample)

```

Waste matches the theoretical waste.

## b
```{r 1_2, echo=FALSE, warning=FALSE}
#code from slides
u <- runif(1000)
y <- 1 - sqrt(1-u)
hist(y)

set.seed(1234)
sample <- c()
for (i in 1:n) {
  # choosing which distribution to sample from with a 50/50 chance
  distribution_multiplier <- sign(runif(1, -1, 1))
  u <- runif(1)
  y <- 1 - sqrt(1-u)
  
  sample <- c(sample, y*distribution_multiplier)

}

hist(sample)

```

## c

```{r 1_3, echo=FALSE, warning=FALSE}

set.seed(12345)
sample <- c()
for (i in 1:n) {
  u1 <- runif(1)
  u2 <- runif(1)
  
  sample <- c(sample, u1 - u2)

}

hist(sample)

```

## d
U1-U2 uses 2 x n random numbers. The inverse cdf also uses 2 x n but performs more calculations. Rejection sampling has some variance, it uses 2 x n + the waste and extra calculations for evaluating the density. U1-U2 is the most efficient.

# Q2

## a


```{r 2_1_1, echo=FALSE, warning=FALSE}


system.time({

n <- 10000000



A <- runif(n, 0, 2*pi)
U <- runif(n)
Y <- sqrt(-2 * log(U))
sample1 <- Y * cos(A)
sample2 <- Y * sin(A)


})

hist(sample1)


```

Box muller gives us a standard normal X ~ N(0, I)
To get $N(\mu, \Sigma)$ we apply $Y = A^TX + \mu$ where A^T comes from the cholesky decomposition of $\Sigma$ 

```{r 2_1_2, echo=FALSE, warning=FALSE}
A_t <- chol(matrix(c(0.6, 0, 0, 0.6), 2, 2))
A_t %*% t(A_t)

```

```{r 2_1_3, echo=FALSE, warning=FALSE}
X <- rbind(sample1, sample2) |> as.matrix()
dim(X)

# 0 mean
Y <- A_t %*% X

hist(Y[1, ])

```
## b
The variance is 0.6 so we take the square root.
```{r 2_2, echo=FALSE, warning=FALSE}
system.time({

x1 <- rnorm(n, 0, sqrt(0.6))
x2 <- rnorm(n, 0, sqrt(0.6))

X <- rbind(x1, x2)
})

```
This was much faster than in a. 

## c
```{r 2_3, echo=FALSE, warning=FALSE}
library(mvtnorm)

normal_bivariate <- function(n) {
  
X <- matrix(0, 2, n)

for (i in 1:n) {
  
  U <- runif(1)
  
  if (U < 0.5) {
      X[, i] <- rmvnorm(1, c(0, 0), diag(2) * 0.6)
  } else {
      X[, i] <- rmvnorm(1, c(1.5, 1.2), diag(2) * 0.5)
  }
  
}

library(ggplot2)

df_main <- data.frame(x = X[1, ], y = X[2, ])
df_points <- data.frame(x = c(0, 1.5), y = c(0, 1.2))

ggplot(df_main, aes(x = x, y = y)) +
  geom_point(alpha = 0.2) + 
  geom_point(data = df_points, aes(x = x, y = y),
             shape = 4,      
             color = "red", 
             size = 5, stroke = 4)
}

set.seed(12345)
normal_bivariate(1000)

```
```{r}
set.seed(12345)
normal_bivariate(10000)
```
Here with a larger n we more clearly see that the density of points drops of further from the means, thanks to the alpha in the plot. We see 2 circles centered around the means which is to be expected.


# Appendix
```{r, ref.label=setdiff(knitr::all_labels(), c('setup')), echo=TRUE, eval=FALSE}

```