---
title: "lab2"
output:
  pdf_document: default
  html_document: default
date: "2026-01-29"
---

## a
```{r}
f <- function(x, y) {
  sin(x+y) + (x-y)^2 - 1.5*x + 2.5*y + 1
  
}

f(3, 2)

x <- y <- seq(-5, 5, length.out = 500)

# outer creates all combinations of x, y
# applying f to the matrix based on x and y
z <- outer(x, y, f)


contour(x = x, y = y, z = z)

```

## b
```{r}
#sin(x+y) + (x-y)^2 - 1.5*x + 2.5*y + 1

g <- function(v) {
  x <- v[1]
  y <- v[2]
  d_dx <- cos(x+y) + 2*x - 2*y - 1.5
  d_dy <- cos(x+y) + 2*y - 2*x + 2.5  
  return( c(d_dx, d_dy) )
    
}

H <- function(v) {
  x <- v[1]
  y <- v[2]
  d_d2x <- -sin(x+y) + 2
  d_d2y <- -sin(x+y) + 2  
  d_dxy <- -sin(x+y) - 2
  
  return(matrix(c(d_d2x, d_dxy, d_dxy, d_d2y), 2, 2))
}

g(c(0,0))
H(c(0,0))
```

## c
Newton is like this for steepest ascent
x_new = x_old - H^-1 * grad

```{r}


Newton <- function(start, eps = 10^(-6), max_it = 50) {
  L2 <- function(v) {
  sqrt(sum(v^2))
  
}


x_t <- start
itterations <- 0
stop_criteria <- T

while (
  stop_criteria & itterations < max_it
) {
  itterations <- itterations + 1
  x_prev <- x_t
  
  
  # For some values the hessian was singular, solving this by adding a small lambda * I to the hessian to get around the issue.
  H_inv <- tryCatch({solve(H(x_prev))}, 
           error = function(x) {solve(H(x_prev) + eps * diag(length(x_t)))})

  x_t <- x_prev - H_inv %*% g(x_prev)
  stop_criteria <- L2(x_t - x_prev) / (eps + L2(x_prev)) > eps
}


eigen_vals <- eigen(H(x_t))$values
cat("Start:", start, ", Itterations:", itterations, ", solution:", x_t, ", function value:", f(x_t[1], x_t[2]), "\neigen_values:", eigen_vals, "\n\n")


}


```

## d

```{r}
Newton(c(-20, -20))
Newton(c(100, -100))
Newton(c(100, 100))
Newton(c(0, 0))
Newton(c(5, 5))
Newton(c(-100, 100))
Newton(c(20, 20))

```
## e
If v^THv < 0 we have a maximum. Loose proof sketch: v^T(Hv) is a dotproduct, Hv is the gradient in that direction. If the gradient, (which points in towards the highest ascent direction), points in the opposite direction as our step, then we are going downwhill. So if all our steps are downhill, negative definite hessian means local maximum. Positive definite means local minimum. If all eigen values are negative we have a local maxima, all are positive means local minimi. Both means saddle point. We see we only found saddle points or local minima in our optimization.

Setting x = y in our function reveals that if y decreases the function will keep decreasing linearly in y. So the function has to global minimum.

# Q 2.

## a

```{r}
# creating the data
x <- c(0, 0, 0, 0.1, 0.1, 0.3, 0.3, 0.9, 0.9, 0.9)
y <- c(0, 0, 1, 0 ,1, 1, 1, 0, 1, 1)

g <- function(v) {
  b0 <- v[1]
  b1 <- v[2]
  
  sum(  y * log( 1 / (1 + exp(-b0 -b1*x)) ) + (1 - y) * log( 1 - ( 1 / (1 + exp(-b0 -b1*x)) ) )  )
  
}

dg1 <- function(v) {
  b0 <- v[1]
  b1 <- v[2]
  
  sum( y - 1 / (1 + exp(-b0 -b1*x)) )
}

dg2 <- function(v) {
  b0 <- v[1]
  b1 <- v[2]
  
  sum( (y - 1 / (1 + exp(-b0 -b1*x))) * x )
}


# The function g to be maximised, partial derivatives dg1, dg2, and gradient
gradient <- function(x)
{
  c(dg1(x), dg2(x))
}

```

```{r}


GD_optimizer <- function(x0, x = c(0, 0, 0, 0.1, 0.1, 0.3, 0.3, 0.9, 0.9, 0.9), y = c(0, 0, 1, 0 ,1, 1, 1, 0, 1, 1),
                         backtracking_shrinkage = 0.5) {
  # Produce a contour plot; define first a grid where function is evaluated
x1grid <- seq(-2, 2.5, by=0.05)
x2grid <- seq(-2, 3, by=0.05)
dx1 <- length(x1grid)
dx2 <- length(x2grid)
dx  <- dx1*dx2
gx  <- matrix(rep(NA, dx), nrow=dx1)
for (i in 1:dx1)
for (j in 1:dx2)
{
  gx[i,j] <- g(c(x1grid[i], x2grid[j]))
}
mgx <- matrix(gx, nrow=dx1, ncol=dx2)
contour(x1grid, x2grid, mgx, nlevels=34)  # Note: For other functions g, you might need to choose another nlevels-value to get a good impression of the function 

#Steepest ascent function:
steepestasc <- function(x0 = x0, eps=1e-6, alpha0=1)
{
  L2 <- function(v) {
  sqrt(sum(v^2))
  
    }
    
  fun_calls <- 0
  grad_calls <- 0
  xt   <- x0
  conv <- 999
  points(xt[1], xt[2], col=2, pch=4, lwd=3)
  while(conv>eps)
  {
    alpha <- alpha0
    
    # xt1 is xtold
    xtold   <- xt
    xt    <- xtold + alpha*gradient(xtold)
    grad_calls <- grad_calls + 1
    
    gxt <- g(xt)
    g_old <- g(xtold)
    fun_calls <- fun_calls + 2
   
    # modified to make less function calls
    while (gxt < g_old)
    {
      
      g_old <- gxt
      
      alpha <- alpha * backtracking_shrinkage
      xt    <- xtold + alpha*gradient(xtold)
      
      gxt <- g(xt)
      fun_calls <- fun_calls + 2
      
    }
    points(xt[1], xt[2], col=2, pch=4, lwd=1)
    
    
    #
    conv <- L2(xt - xtold)
  }
  points(xt[1], xt[2], col=4, pch=4, lwd=3)
  cat("Function calls:", fun_calls, "\n")
  cat("Gradient calls:", grad_calls, "\n")

  return(xt)
}

steepestasc(c(-0.2, 1))

}



GD_optimizer(c(-0.2, 1), backtracking_shrinkage = 0.7)

GD_optimizer(c(-0.2, 1), backtracking_shrinkage = 0.3)

GD_optimizer(c(-0.2, 1), backtracking_shrinkage = 0.5)

```
We set the change beteween xt and xtold to be smaller than 10^-6 so we can trust 5 decimals.
With different backtracking methods we get way more function calls if the shrinkage factor for alpha is larger there needs to be more calls made. 0.3 and 0.5 perform similarly. 

## c
```{r}

# optim minimizes so we try to maximize by multiplying by -1 via fnscale
optim(c(-0.2, 1), fn = g, gr = gradient, method = "BFGS", control = list(fnscale = -1))

optim(c(-0.2, 1), fn = g, method = "Nelder-Mead", control = list(fnscale = -1))


```
They converge way faster and therefore use fewer function-calls. Nelson-meader converging faster is interesting as it has no theoretical convergence speed. 

our result from b: -0.00935(9121)  1.26282(1823) precise to the 5th decimal. BFGS appears to have 4 decimals of precision and nelson meader appears to have 3. Digging into the documentation we find that the relative tolerance is 1e-8, so the amount of decimals depends on the size of the root.

## d
```{r}

model <- glm(y ~ x, family = "binomial")

a <- summary(model)
a$coefficients[, 1]

```

It matches well with our optimized results for the first 5 decimals.