---
title: "Lab 2"
author: "Aron Enström (aroen488), Sergey Volchkov (servo519)"
date: "2026-01-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1


### 1.1. Function plot
```{r 1_1, echo=FALSE}
#1.1
f <- function(x, y) sin(x + y) + (x - y)^2 - 1.5*x + 2.5*y + 1

x <- seq(-1.5, 4, length.out = 300)
y <- seq(-3,   4, length.out = 300)

z <- outer(x, y, Vectorize(f))

filled.contour(x, y, z,
  nlevels = 30,
  color.palette = hcl.colors,
  plot.axes = {
    axis(1); axis(2)
    contour(x, y, z, add = TRUE, drawlabels = TRUE)
  }
)
```

### 1.2. Gradient and Hessian

For 
$$
f(x,y)=\sin(x+y)+(x-y)^2-1.5x+2.5y+1,
$$
the gradient is
$$
\nabla f(x,y)=
\begin{pmatrix}
\frac{\partial f}{\partial x}\\[4pt]
\frac{\partial f}{\partial y}
\end{pmatrix}
=
\begin{pmatrix}
\cos(x+y)+2(x-y)-1.5\\[4pt]
\cos(x+y)-2(x-y)+2.5
\end{pmatrix}.
$$

The Hessian matrix (matrix of second partial derivatives) is
$$
\nabla^2 f(x,y)=
\begin{pmatrix}
\frac{\partial^2 f}{\partial x^2} & \frac{\partial^2 f}{\partial x\partial y}\\[4pt]
\frac{\partial^2 f}{\partial y\partial x} & \frac{\partial^2 f}{\partial y^2}
\end{pmatrix}
=
\begin{pmatrix}
-\sin(x+y)+2 & -\sin(x+y)-2\\[4pt]
-\sin(x+y)-2 & -\sin(x+y)+2
\end{pmatrix},
$$
where $\frac{\partial^2 f}{\partial x\partial y}=\frac{\partial^2 f}{\partial y\partial x}$.


```{r 1_2, echo=FALSE}
#1.2
## Gradient
grad_f <- function(x, y) {
  c(
    dx = cos(x + y) + 2*(x - y) - 1.5,
    dy = cos(x + y) - 2*(x - y) + 2.5
  )
}

## Hessian
hess_f <- function(x, y) {
  s <- -sin(x + y)
  matrix(
    c(s + 2, s - 2,
      s - 2, s + 2),
    nrow = 2, byrow = TRUE,
    dimnames = list(c("x", "y"), c("x", "y"))
  )
}

## wrappers that take par = c(x, y) for optimization
fn_par   <- function(par) f(par[1], par[2])
gr_par   <- function(par) grad_f(par[1], par[2])
hess_par <- function(par) hess_f(par[1], par[2])

## test
# par0 <- c(0, 0)
# fn_par(par0)
# gr_par(par0)
# hess_par(par0)
```

### 1.3 Newton algorithm

We are asked to search for the global minimum solution within the box 

$$
x \in [-1.5,\,4], \qquad y \in [-3,\,4].
$$


so the algorithm code needs to be adjusted to prevent the search path wandering outside this box. 

Compared to the original (unconstrained, “pure”) Newton algorithm, we introduced three standard safeguards and one additional fallback so the method behaves well on a bounded, potentially nonconvex objective.

First, we enforced the box constraints by *projecting* every proposed iterate back into the feasible region: after forming a candidate update, we replace it by its componentwise clipped version so that $x\in[-1.5,4]$ and $y\in[-3,4]$ always hold.

Second, we stabilized the Newton step computation by *regularizing* the Hessian. Instead of solving $H(x_k,y_k)\,p_k=\nabla f(x_k,y_k)$ (which can fail when $H$ is singular or ill-conditioned), we solve $(H(x_k,y_k)+\lambda I)\,p_k=\nabla f(x_k,y_k)$ and increase $\lambda$ as needed until the linear system is solvable.

Third, we added a *backtracking line search* (step damping). Rather than always taking a full Newton step (i.e., step size $\alpha=1$), we update via $(x_{k+1},y_{k+1})=P\big((x_k,y_k)-\alpha p_k\big)$ (where $P(\cdot)$ denotes the projection operator) and reduce $\alpha$ (e.g., $\alpha\leftarrow\beta\alpha$ with $0<\beta<1$) until an Armijo-type sufficient decrease condition is satisfied, which prevents overshooting and improves robustness on nonconvex landscapes.

Finally, to avoid stalling at the boundary (where projection can collapse the effective Newton step and force $\alpha$ to become extremely small), we included a fallback to a *projected-gradient* step when the Newton step fails to produce progress. Concretely, we switch to $(x_{k+1},y_{k+1})=P\big((x_k,y_k)-\alpha\nabla f(x_k,y_k)\big)$, which provides a feasible descent direction and helps the algorithm move along edges instead of getting trapped at corners.



```{r 1_3, echo=FALSE}
#1.3

newton2d_box <- function(fn, gr, hess, x0, y0,
                                lower, upper,
                                tol = 1e-8, maxit = 200,
                                lambda0 = 1e-6, lambda_mult = 10,
                                alpha0 = 1, beta = 0.5, c1 = 1e-4,
                                pg_alpha0 = 1, pg_beta = 0.5,
                                eps_active = 1e-10,
                                ..., verbose = FALSE, keep_path = TRUE) {

  project <- function(par) pmin(pmax(par, lower), upper)

  ## projected-gradient infinity norm (KKT-style for box constraints)
  projgrad_inf <- function(x, g, eps = 1e-12) {
    pg <- g
    for (i in 1:2) {
      if (x[i] <= lower[i] + eps) pg[i] <- min(g[i], 0)
      if (x[i] >= upper[i] - eps) pg[i] <- max(g[i], 0)
    }
    max(abs(pg))
  }

  ## "gradient mapping" direction: d = P(x - a g) - x (feasible descent step)
  gradmap_dir <- function(x, g, a) project(x - a * g) - x

  x <- project(c(x0, y0))
  f0 <- fn(x[1], x[2], ...)

  path <- NULL
  if (keep_path) {
    path <- data.frame(iter = 0L, x = x[1], y = x[2], f = f0,
                       step = NA_real_, method = "init",
                       projgrad = projgrad_inf(x, gr(x[1], x[2], ...)))
  }

  for (k in 1:maxit) {
    g <- gr(x[1], x[2], ...)
    pg_norm <- projgrad_inf(x, g)

    if (pg_norm < tol) {
      return(list(par = x, value = f0, iter = k - 1, converged = TRUE,
                  projgrad = pg_norm, path = path))
    }

    ## ---- Attempt Newton step (with regularization + projected Armijo) ----
    H <- hess(x[1], x[2], ...)
    lambda <- lambda0
    p <- NULL
    repeat {
      p_try <- tryCatch(solve(H + lambda * diag(2), g), error = function(e) NULL)
      if (!is.null(p_try)) { p <- p_try; break }
      lambda <- lambda * lambda_mult
      if (lambda > 1e12) { p <- NULL; break }
    }

    newton_ok <- FALSE
    a <- alpha0
    if (!is.null(p)) {
      repeat {
        x_new <- project(x - a * p)
        f_new <- fn(x_new[1], x_new[2], ...)
        armijo_rhs <- f0 - c1 * a * sum(g * p)  # sufficient decrease
        if (is.finite(f_new) && f_new <= armijo_rhs && any(abs(x_new - x) > 0)) {
          newton_ok <- TRUE
          break
        }
        a <- a * beta
        if (a < 1e-12) break
      }
    }

    if (newton_ok) {
      x <- x_new; f0 <- f_new
      if (keep_path) path <- rbind(path, data.frame(iter = k, x = x[1], y = x[2], f = f0,
                                                   step = a, method = "Newton",
                                                   projgrad = pg_norm))
      if (verbose) cat(sprintf("iter=%d Newton  x=%.8f y=%.8f f=%.10f step=%.1e ||pg||_inf=%.3e\n",
                               k, x[1], x[2], f0, a, pg_norm))
      next
    }

    ## ---- Fallback: projected-gradient step (guarantees feasible movement) ----
    a_pg <- pg_alpha0
    repeat {
      d <- gradmap_dir(x, g, a_pg)
      x_new <- x + d
      f_new <- fn(x_new[1], x_new[2], ...)
      if (is.finite(f_new) && f_new <= f0) break  # simple monotone decrease
      a_pg <- a_pg * pg_beta
      if (a_pg < 1e-12) break
    }

    ## If even PG can’t decrease, stop (we’re at a constrained stationary-ish point)
    if (sqrt(sum((x_new - x)^2)) < tol) {
      return(list(par = x, value = f0, iter = k, converged = FALSE,
                  projgrad = pg_norm, path = path,
                  message = "No progress (Newton+PG failed to decrease)."))
    }

    x <- x_new; f0 <- f_new
    if (keep_path) path <- rbind(path, data.frame(iter = k, x = x[1], y = x[2], f = f0,
                                                 step = a_pg, method = "PG",
                                                 projgrad = pg_norm))
    if (verbose) cat(sprintf("iter=%d PG      x=%.8f y=%.8f f=%.10f step=%.1e ||pg||_inf=%.3e\n",
                             k, x[1], x[2], f0, a_pg, pg_norm))
  }

  list(par = x, value = f0, iter = maxit, converged = FALSE,
       projgrad = projgrad_inf(x, gr(x[1], x[2], ...)),
       path = path)
}

# newton2d_box <- function(fn, gr, hess, x0, y0,
#                                 lower, upper,
#                                 tol = 1e-8, maxit = 100,
#                                 lambda0 = 1e-6, lambda_mult = 10,
#                                 alpha0 = 1, beta = 0.5, c1 = 1e-4,
#                                 eps_active = 1e-10,
#                                 ..., verbose = FALSE, keep_path = TRUE) {
# 
#   project <- function(par) pmin(pmax(par, lower), upper)
# 
#   projgrad_inf <- function(x, g, eps = 1e-12) {
#     pg <- g
#     for (i in 1:2) {
#       if (x[i] <= lower[i] + eps) pg[i] <- min(g[i], 0)
#       if (x[i] >= upper[i] - eps) pg[i] <- max(g[i], 0)
#     }
#     max(abs(pg))
#   }
# 
#   ## Active-set: free if not too close to either bound
#   free_set <- function(x) {
#     c(
#       (x[1] > lower[1] + eps_active) && (x[1] < upper[1] - eps_active),
#       (x[2] > lower[2] + eps_active) && (x[2] < upper[2] - eps_active)
#     )
#   }
# 
#   x <- project(c(x0, y0))
#   f0 <- fn(x[1], x[2], ...)
# 
#   path <- NULL
#   if (keep_path) {
#     path <- data.frame(iter = 0L, x = x[1], y = x[2], f = f0,
#                        step = NA_real_, projgrad = NA_real_, lambda = NA_real_,
#                        free_x = NA, free_y = NA)
#   }
# 
#   for (k in 1:maxit) {
#     g <- gr(x[1], x[2], ...)
#     H <- hess(x[1], x[2], ...)
#     pg_norm <- projgrad_inf(x, g)
# 
#     free <- free_set(x)
#     if (keep_path) {
#       path <- rbind(path, data.frame(iter = k, x = x[1], y = x[2], f = f0,
#                                      step = NA_real_, projgrad = pg_norm, lambda = NA_real_,
#                                      free_x = free[1], free_y = free[2]))
#     }
# 
#     if (pg_norm < tol) {
#       return(list(par = x, value = f0, iter = k - 1, converged = TRUE,
#                   projgrad = pg_norm, path = path))
#     }
# 
#     ## If nothing is free, we're at a corner/edge; stop (or switch to PG)
#     if (!any(free)) {
#       return(list(par = x, value = f0, iter = k - 1, converged = FALSE,
#                   projgrad = pg_norm, path = path,
#                   message = "Stuck at corner: no free variables."))
#     }
# 
#     ## Reduced Newton direction p: only update free components
#     p_full <- c(0, 0)
#     idx <- which(free)
# 
#     lambda <- lambda0
#     repeat {
#       Hred <- H[idx, idx, drop = FALSE]
#       gred <- g[idx]
#       p_try <- tryCatch(solve(Hred + lambda * diag(length(idx)), gred),
#                         error = function(e) NULL)
#       if (!is.null(p_try)) { p_full[idx] <- p_try; break }
#       lambda <- lambda * lambda_mult
#       if (lambda > 1e12) stop("Could not compute reduced Newton step.")
#     }
# 
#     ## Line search on reduced step (projection just guards numerics)
#     a <- alpha0
#     repeat {
#       x_new <- project(x - a * p_full)
#       f_new <- fn(x_new[1], x_new[2], ...)
# 
#       armijo_rhs <- f0 - c1 * a * sum(g[idx] * p_full[idx])  # uses free components
#       if (is.finite(f_new) && f_new <= armijo_rhs) break
# 
#       a <- a * beta
#       if (a < 1e-12) break
#     }
# 
#     if (verbose) {
#       cat(sprintf("iter=%d  x=%.8f y=%.8f  f=%.10f  step=%.1e  ||pg||_inf=%.3e  lambda=%.1e  free=(%d,%d)\n",
#                   k, x_new[1], x_new[2], f_new, a, pg_norm, lambda, free[1], free[2]))
#     }
# 
#     if (keep_path) {
#       path[nrow(path), "step"] <- a
#       path[nrow(path), "lambda"] <- lambda
#     }
# 
#     ## Secondary stop: tiny move AND small projected grad
#     if (sqrt(sum((x_new - x)^2)) < tol && pg_norm < tol) {
#       return(list(par = x_new, value = f_new, iter = k, converged = TRUE,
#                   projgrad = pg_norm, path = path))
#     }
# 
#     ## If the step collapsed, bail out instead of looping forever
#     if (a < 1e-12) {
#       return(list(par = x, value = f0, iter = k, converged = FALSE,
#                   projgrad = pg_norm, path = path,
#                   message = "Line search failed (step underflow)."))
#     }
# 
#     x <- x_new
#     f0 <- f_new
#   }
# 
#   list(par = x, value = f0, iter = maxit, converged = FALSE,
#        projgrad = projgrad_inf(x, gr(x[1], x[2], ...)),
#        path = path)
# }

# newton2d_box <- function(fn, gr, hess, x0, y0,
#                          lower, upper,
#                          tol = 1e-8, maxit = 100,
#                          lambda0 = 1e-6, lambda_mult = 10,
#                          alpha0 = 1, beta = 0.5, c1 = 1e-4,
#                          ..., verbose = FALSE, keep_path = TRUE) {
# 
#   project <- function(par) pmin(pmax(par, lower), upper)
# 
#   ## infinity-norm of projected gradient (KKT-style for box constraints)
#   projgrad_inf <- function(x, g, eps = 1e-12) {
#     pg <- g
#     for (i in 1:2) {
#       if (x[i] <= lower[i] + eps) pg[i] <- min(g[i], 0)
#       if (x[i] >= upper[i] - eps) pg[i] <- max(g[i], 0)
#     }
#     max(abs(pg))
#   }
# 
#   x <- project(c(x0, y0))
#   f0 <- fn(x[1], x[2], ...)
# 
#   ## store path (x, y, f, step, projgrad, lambda)
#   path <- NULL
#   if (keep_path) {
#     path <- data.frame(iter = 0L, x = x[1], y = x[2], f = f0,
#                        step = NA_real_, projgrad = NA_real_, lambda = NA_real_)
#   }
# 
#   for (k in 1:maxit) {
#     g <- gr(x[1], x[2], ...)
#     pg_norm <- projgrad_inf(x, g)
# 
#     if (keep_path) path <- rbind(path, data.frame(
#       iter = k, x = x[1], y = x[2], f = f0,
#       step = NA_real_, projgrad = pg_norm, lambda = NA_real_
#     ))
# 
#     if (pg_norm < tol) {
#       return(list(par = x, value = f0, iter = k - 1, converged = TRUE,
#                   projgrad = pg_norm, path = path))
#     }
# 
#     H <- hess(x[1], x[2], ...)
#     I2 <- diag(2)
# 
#     ## Regularize Hessian until solvable
#     lambda <- lambda0
#     p <- NULL
#     repeat {
#       p_try <- tryCatch(solve(H + lambda * I2, g), error = function(e) NULL)
#       if (!is.null(p_try)) { p <- p_try; break }
#       lambda <- lambda * lambda_mult
#       if (lambda > 1e12) stop("Could not compute Newton step (even after regularization).")
#     }
# 
#     ## Backtracking line search with Armijo sufficient decrease
#     a <- alpha0
#     repeat {
#       x_new <- project(x - a * p)
#       f_new <- fn(x_new[1], x_new[2], ...)
# 
#       armijo_rhs <- f0 - c1 * a * sum(g * p)
# 
#       if (is.finite(f_new) && f_new <= armijo_rhs) break
#       a <- a * beta
#       if (a < 1e-12) break
#     }
# 
#     if (verbose) {
#       cat(sprintf(
#         "iter=%d  x=%.8f y=%.8f  f=%.10f  step=%.1e  ||pg||_inf=%.3e  lambda=%.1e\n",
#         k, x_new[1], x_new[2], f_new, a, pg_norm, lambda
#       ))
#     }
# 
#     if (keep_path) {
#       path[nrow(path), "step"] <- a
#       path[nrow(path), "lambda"] <- lambda
#       path[nrow(path), "projgrad"] <- pg_norm
#     }
# 
#     ## Only use step-size convergence as a secondary criterion
#     if (sqrt(sum((x_new - x)^2)) < tol && pg_norm < tol) {
#       return(list(par = x_new, value = f_new, iter = k, converged = TRUE,
#                   projgrad = pg_norm, path = path))
#     }
# 
#     x <- x_new
#     f0 <- f_new
#   }
# 
#   list(par = x, value = f0, iter = maxit, converged = FALSE,
#        projgrad = projgrad_inf(x, gr(x[1], x[2], ...)),
#        path = path)
# }

# newton2d_box <- function(fn, gr, hess, x0, y0,
#                          lower, upper,
#                          tol = 1e-8, maxit = 100,
#                          lambda0 = 1e-6, lambda_mult = 10,
#                          alpha0 = 1, beta = 0.5, c1 = 1e-4,
#                          ..., verbose = FALSE) {
# 
#   project <- function(par) pmin(pmax(par, lower), upper)
# 
#   ## infinity-norm of projected gradient (KKT-style for box constraints)
#   projgrad_inf <- function(x, g, eps = 1e-12) {
#     pg <- g
#     for (i in 1:2) {
#       if (x[i] <= lower[i] + eps) pg[i] <- min(g[i], 0)  # should be >=0 at lower bound
#       if (x[i] >= upper[i] - eps) pg[i] <- max(g[i], 0)  # should be <=0 at upper bound 
#     }
#     max(abs(pg))
#   }
# 
#   x <- project(c(x0, y0))
#   f0 <- fn(x[1], x[2], ...)
# 
#   for (k in 1:maxit) {
#     g <- gr(x[1], x[2], ...)
#     pg_norm <- projgrad_inf(x, g)
# 
#     if (pg_norm < tol) {
#       return(list(par = x, value = f0, iter = k - 1, converged = TRUE,
#                   projgrad = pg_norm))
#     }
# 
#     H <- hess(x[1], x[2], ...)
#     I2 <- diag(2)
# 
#     ## Regularize Hessian until solvable
#     lambda <- lambda0
#     p <- NULL
#     repeat {
#       p_try <- tryCatch(solve(H + lambda * I2, g), error = function(e) NULL)
#       if (!is.null(p_try)) { p <- p_try; break }
#       lambda <- lambda * lambda_mult
#       if (lambda > 1e12) stop("Could not compute Newton step (even after regularization).")
#     }
# 
#     ## Backtracking line search with Armijo sufficient decrease
#     a <- alpha0
#     repeat {
#       x_new <- project(x - a * p)
#       f_new <- fn(x_new[1], x_new[2], ...)
# 
#       ## directional derivative along -p is approx -g^T p
#       armijo_rhs <- f0 - c1 * a * sum(g * p)
# 
#       if (is.finite(f_new) && f_new <= armijo_rhs) break
#       a <- a * beta
#       if (a < 1e-12) break
#     }
# 
#     if (verbose) {
#       cat(sprintf(
#         "iter=%d  x=%.8f y=%.8f  f=%.10f  step=%.1e  ||pg||_inf=%.3e  lambda=%.1e\n",
#         k, x_new[1], x_new[2], f_new, a, pg_norm, lambda
#       ))
#     }
# 
#     ## Only use step-size convergence as a secondary criterion
#     if (sqrt(sum((x_new - x)^2)) < tol && pg_norm < tol) {
#       return(list(par = x_new, value = f_new, iter = k, converged = TRUE,
#                   projgrad = pg_norm))
#     }
# 
#     x <- x_new
#     f0 <- f_new
#   }
# 
#   list(par = x, value = f0, iter = maxit, converged = FALSE,
#        projgrad = projgrad_inf(x, gr(x[1], x[2], ...)))
# }


# newton2d_box <- function(fn, gr, hess, x0, y0,
#                          lower, upper,
#                          tol = 1e-8, maxit = 100,
#                          lambda0 = 1e-6, lambda_mult = 10,
#                          alpha0 = 1, beta = 0.5,
#                          ..., verbose = FALSE) {
# 
#   #To keep the results within the prescribed x,y limits
#   project <- function(par) pmin(pmax(par, lower), upper)
# 
#   x <- project(c(x0, y0))
#   f0 <- fn(x[1], x[2], ...)
# 
#   for (k in 1:maxit) {
#     g <- gr(x[1], x[2], ...)
#     H <- hess(x[1], x[2], ...)
#     I2 <- diag(2)
# 
#     ## regularize until solvable
#     lambda <- lambda0
#     p <- NULL
#     repeat {
#       p_try <- tryCatch(solve(H + lambda * I2, g), error = function(e) NULL)
#       if (!is.null(p_try)) { p <- p_try; break }
#       lambda <- lambda * lambda_mult
#       if (lambda > 1e12) stop("Could not compute Newton step.")
#     }
# 
#     ## backtracking line search with projection
#     a <- alpha0
#     repeat {
#       x_new <- project(x - a * p)
#       f_new <- fn(x_new[1], x_new[2], ...)
#       if (is.finite(f_new) && f_new <= f0) break
#       a <- a * beta
#       if (a < 1e-12) break
#     }
# 
#     if (verbose) {
#       cat(sprintf("iter=%d  x=%.8f y=%.8f  f=%.10f  step=%.1e\n",
#                   k, x_new[1], x_new[2], f_new, a))
#     }
# 
#     if (sqrt(sum((x_new - x)^2)) < tol) {
#       return(list(par = x_new, value = f_new, iter = k, converged = TRUE))
#     }
# 
#     x <- x_new
#     f0 <- f_new
#   }
# 
#   list(par = x, value = f0, iter = maxit, converged = FALSE)
# }

# ## x_{k+1} = x_k - H(x_k)^{-1} * g(x_k)
# newton2d_reg <- function(fn, gr, hess, x0, y0,
#                          tol = 1e-8, maxit = 100,
#                          lambda0 = 1e-6, lambda_mult = 10,
#                          ..., verbose = FALSE) {
# 
#   x <- c(x0, y0)
# 
#   for (k in 1:maxit) {
#     g <- gr(x[1], x[2], ...)
#     H <- hess(x[1], x[2], ...)
#     I2 <- diag(2)
# 
#     lambda <- lambda0
#     p <- NULL
# 
#     ## Keep increasing lambda until the linear system is solvable
#     repeat {
#       Hp <- H + lambda * I2
#       p_try <- tryCatch(solve(Hp, g), error = function(e) NULL)  # catch singularity
#       if (!is.null(p_try)) { p <- p_try; break }
#       lambda <- lambda * lambda_mult
#       if (lambda > 1e12) stop("Hessian remained singular/ill-conditioned even after regularization.")
#     }
# 
#     x_new <- x - p
# 
#     if (verbose) {
#       cat(sprintf("iter=%d  x=%.10f y=%.10f  f=%.10f  ||grad||=%.3e  lambda=%.1e\n",
#                   k, x_new[1], x_new[2], fn(x_new[1], x_new[2], ...),
#                   sqrt(sum(g^2)), lambda))
#     }
# 
#     if (sqrt(sum((x_new - x)^2)) < tol) {
#       return(list(par = x_new, value = fn(x_new[1], x_new[2], ...),
#                   iter = k, converged = TRUE))
#     }
#     x <- x_new
#   }
# 
#   list(par = x, value = fn(x[1], x[2], ...), iter = maxit, converged = FALSE)
# }
# 
lower <- c(-1.5, -3)
upper <- c( 4.0,  4)
# res <- newton2d_box(fn = f, gr = grad_f, hess = hess_f, 
#                     x0 = 0, y0 = 0, lower=lower, upper=upper, verbose = TRUE)
# res$par
# res$value
# 
# 
res00 <- newton2d_box(fn = f, gr = grad_f, hess = hess_f,
                      x0 = 0, y0 = 0,
                      lower = lower, upper = upper,
                      verbose = FALSE, keep_path = TRUE)


# head(res00$path)

filled.contour(x, y, z,
  nlevels = 30,
  color.palette = hcl.colors,
  plot.title = title(main = "Testing convergence from starting point at (0,0)",
                     xlab = "x", ylab = "y"),
  plot.axes = {
    axis(1); axis(2)
    contour(x, y, z, add = TRUE, drawlabels = TRUE)  

    ## Newton path: line + points (start = open circle, end = filled circle)
    lines(res00$path$x, res00$path$y, col = "red", lwd = 2)
    points(res00$path$x, res00$path$y, col = "red", pch = 1, cex = 0.9)
    points(res00$path$x[1], res00$path$y[1], col = "red", pch = 21, bg = "white", cex = 1.3)
    points(tail(res00$path$x, 1), tail(res00$path$y, 1), col = "red", pch = 19, cex = 1.3)

    legend("topright",
           legend = c("Newton path", "Start", "End"),
           col = c("red", "red", "red"),
           lwd = c(2, NA, NA),
           pch = c(NA, 21, 19),
           pt.bg = c(NA, "white", NA),
           bty = "n")
  }
)
```

Comparing with the results produced by L-BFGS-B algorithm (BGFS with constraints) on a small sample of points.

```{r 1_3_1, echo=FALSE}
lower <- c(-1.5, -3)
upper <- c( 4.0,  4)

one_run <- function(x0, y0) {
  nres <- newton2d_box(f, grad_f, hess_f, x0, y0, lower, upper)

  ores <- optim(
    par = c(x0, y0),
    fn = fn_par, gr = gr_par,
    method = "L-BFGS-B",
    lower = lower, upper = upper
  )  # box constraints supported by L-BFGS-B

  data.frame(
    x0 = x0, y0 = y0,
    newton_iter = nres$iter,
    newton_x = nres$par[1], newton_y = nres$par[2],
    newton_value = nres$value,
    optim_iter = ores$counts[["function"]],
    optim_x = ores$par[1], optim_y = ores$par[2],
    optim_value = ores$value
  )
}


starts <- rbind(
  c(-1.0, -2.0),
  c( 3.5,  3.5),
  c( 0.0,  0.0),
  c( 4.0, -3.0),
  c(-1.5,  4.0)
)
colnames(starts) <- c("x0", "y0")

res_table <- do.call(
  rbind,
  lapply(seq_len(nrow(starts)), function(i) one_run(starts[i, 1], starts[i, 2]))
)

res_table_fmt <- res_table
num_cols <- names(res_table_fmt)[sapply(res_table_fmt, is.numeric)]

res_table_fmt[num_cols] <- lapply(res_table_fmt[num_cols], \(v) formatC(v, format = "f", digits = 6))
# print(res_table_fmt, row.names = FALSE)

## distinct solutions reached (rounded to group near-equal points)
# unique(round(res_table[, c("newton_x", "newton_y")], 6))
# unique(round(res_table[, c("optim_x", "optim_y")], 6))


filled.contour(x, y, z,
  nlevels = 30,
  color.palette = hcl.colors,
  plot.title = title(main = "Comparing the results with L-BFGS-B algorithm\non a sample of points",
                     xlab = "x", ylab = "y"),
  plot.axes = {
    axis(1); axis(2)
    contour(x, y, z, add = TRUE, drawlabels = TRUE)

    ## Overlay solutions
    points(res_table$newton_x, res_table$newton_y,
           pch = 19, col = "red", cex = 2.1)
    points(res_table$optim_x,  res_table$optim_y,
           pch = 19, col = "dodgerblue3", cex = 1.1)

    legend("topright",
           legend = c("Newton (ours)", "optim L-BFGS-B"),
           col = c("red", "dodgerblue3"),
           pch = 19, bty = "n")
  }
)
```

We can see from the above diagram that our algorithm converges in the same spot from all the randomly selected starting points, the endpoint seems to be where we expect the global minimum to lie on our heatmap plot and it coincides with the one produced by the L-BFGS-B algorithm which gives us confidence that it is working correctly.

### 1.4 Testing with different starting points

To identify the points from which the algorithms does not converge to the true minimum, we need to use a denser grid:

```{r 1_4_1, echo=FALSE}
#1.4
lower <- c(-1.5, -3)
upper <- c( 4.0,  4)

## multi-start L-BFGS-B to approximate the global minimum in the box
x_st <- seq(lower[1], upper[1], length.out = 15)
y_st <- seq(lower[2], upper[2], length.out = 15)
starts_ref <- expand.grid(x0 = x_st, y0 = y_st)  # all combinations

run_lbfgsb <- function(x0, y0) {
  o <- optim(par = c(x0, y0),
             fn = fn_par, gr = gr_par,
             method = "L-BFGS-B",
             lower = lower, upper = upper)  # bound constraints
  list(par = o$par, value = o$value, counts = o$counts)
}

ref_list <- lapply(seq_len(nrow(starts_ref)), function(i) run_lbfgsb(starts_ref$x0[i], starts_ref$y0[i]))
ref_vals <- sapply(ref_list, `[[`, "value")
best_i <- which.min(ref_vals)

x_star_ref <- ref_list[[best_i]]$par
f_star_ref <- ref_list[[best_i]]$value
# x_star_ref; f_star_ref

tol_val <- 1e-6

## grid (or increase resolution / use random points)
xg <- seq(lower[1], upper[1], length.out = 41)
yg <- seq(lower[2], upper[2], length.out = 41)
starts <- expand.grid(x0 = xg, y0 = yg) 

one_run_newton <- function(x0, y0) {
  nres <- newton2d_box(f, grad_f, hess_f, x0, y0, lower, upper)

  data.frame(
    x0 = x0, y0 = y0,
    converged = isTRUE(nres$converged),
    iters = nres$iter,
    x = nres$par[1], y = nres$par[2],
    f = nres$value,
    miss_true_min = (!isTRUE(nres$converged)) || (nres$value > f_star_ref + tol_val)
  )
}

scan_res <- do.call(rbind, lapply(seq_len(nrow(starts)), function(i) one_run_newton(starts$x0[i], starts$y0[i])))

## starting points where Newton does NOT reach the (estimated) global min
bad_starts <- subset(scan_res, miss_true_min)

# nrow(bad_starts)
# head(bad_starts)

cols <- ifelse(scan_res$miss_true_min, "red", "dodgerblue3")
plot(scan_res$x0, scan_res$y0, col = cols, pch = 15, cex = 0.6,
     xlab = "x0", ylab = "y0",
     main = "Starts: Newton misses (red) vs reaches (blue) the true minimum")
points(x_star_ref[1], x_star_ref[2], pch = 8, cex = 1.5, lwd = 2)
legend("topright", legend = c("Reaches true min", "Misses true min", "Reference min"),
       col = c("dodgerblue3", "red", "black"), pch = c(15, 15, 8), bty = "n")
```

Using the above output to see which points the algorithm converges to when it does not converge to the true minimum:

```{r 1_4_2, echo=FALSE}

endpts <- do.call(rbind, lapply(seq_len(nrow(bad_starts)), function(i) {
  x0 <- bad_starts$x0[i]
  y0 <- bad_starts$y0[i]

  r <- newton2d_box(f, grad_f, hess_f, x0, y0, lower, upper, keep_path = TRUE)

  data.frame(
    x0 = x0, y0 = y0,
    end_x = r$par[1],
    end_y = r$par[2],
    end_f = r$value,
    converged = isTRUE(r$converged),
    iter = r$iter
  )
}))

endpts<-unique(round(endpts[, c("end_x","end_y","end_f", "converged")], 6))

filled.contour(x, y, z,
  nlevels = 30,
  color.palette = hcl.colors,
  plot.title = title(main = "Attractors other than the global minimum",
                     xlab = "x", ylab = "y"),
  plot.axes = {
    axis(1); axis(2)
    contour(x, y, z, add = TRUE, drawlabels = TRUE)

    ## Endpoints of Newton from the bad starts
    cols <- ifelse(endpts$converged, "red", "orange")
    points(endpts$end_x, endpts$end_y, pch = 19, col = cols, cex = 1.1)

    ## Global minimum (reference)
    points(x_star_ref[1], x_star_ref[2], pch = 8, col = "white", cex = 1.6, lwd = 2)

    legend("topright",
           legend = c("Newton endpoints (converged)",
                      "Newton endpoints (not converged)",
                      "Global minimum (reference)"),
           pch = c(19, 19, 8),
           col = c("red", "orange", "white"),
           pt.cex = c(1.1, 1.1, 1.3),
           text.col = "white",
           bty = "n")
  }
)
```

### 1.5 Understanding the nature of the incorrect endpoints

```{r 1_5, echo=FALSE}
#1.4
# Since none of the detected endpoints lies near the boundary, we can assume it is not
# due to the boundary interfering with the workings of our algorithm
# We can therefore classify the endpoints by eigenvalues of H(x,y)
# - all eigenvalues >0 -> local minimum
# - mixed signs -> saddle
# - all <0 -> local maximum
classify_interior <- function(x, y, grad_f, hess_f, tol_g = 1e-6) {
  g <- grad_f(x, y)
  H <- hess_f(x, y)
  eg <- eigen(H, symmetric = TRUE)$values

  # if (sqrt(sum(g^2)) > tol_g) return("Not stationary (interior)")

  if (all(eg > 0)) "Local min"
  else if (all(eg < 0)) "Local max"
  else "Saddle / indefinite"
}

endpts$class <- apply(endpts[, c("end_x", "end_y")], 1, \(v) {
  classify_interior(v[1], v[2], grad_f, hess_f)
})

global_row <- data.frame(
  end_x = x_star_ref[1],
  end_y = x_star_ref[2],
  end_f = f(x_star_ref[1], x_star_ref[2]),
  class = "Global minimum",
  converged = TRUE
)

endpts2 <- rbind(endpts, global_row)

table(endpts2$class)

endpts2[order(endpts2$end_f), ]
```


## 2


## Appendix


```{r, ref.label=setdiff(knitr::all_labels(), c('setup')), echo=TRUE, eval=FALSE}



```
