---
title: "Lab 2"
author: "Aron Enström (aroen488), Sergey Volchkov (servo519)"
date: "2026-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1


### 1.1. Function plot
```{r 1_1, echo=FALSE}
#1.1
f <- function(x, y) sin(x + y) + (x - y)^2 - 1.5*x + 2.5*y + 1

x <- seq(-1.5, 4, length.out = 300)
y <- seq(-3,   4, length.out = 300)

z <- outer(x, y, Vectorize(f))

filled.contour(x, y, z,
  nlevels = 30,
  color.palette = hcl.colors,
  plot.axes = {
    axis(1); axis(2)
    contour(x, y, z, add = TRUE, drawlabels = TRUE)
  }
)
```

### 1.2. Gradient and Hessian

For 
$$
f(x,y)=\sin(x+y)+(x-y)^2-1.5x+2.5y+1,
$$
the gradient is
$$
\nabla f(x,y)=
\begin{pmatrix}
\frac{\partial f}{\partial x}\\[4pt]
\frac{\partial f}{\partial y}
\end{pmatrix}
=
\begin{pmatrix}
\cos(x+y)+2(x-y)-1.5\\[4pt]
\cos(x+y)-2(x-y)+2.5
\end{pmatrix}.
$$

The Hessian matrix (matrix of second partial derivatives) is
$$
\nabla^2 f(x,y)=
\begin{pmatrix}
\frac{\partial^2 f}{\partial x^2} & \frac{\partial^2 f}{\partial x\partial y}\\[4pt]
\frac{\partial^2 f}{\partial y\partial x} & \frac{\partial^2 f}{\partial y^2}
\end{pmatrix}
=
\begin{pmatrix}
-\sin(x+y)+2 & -\sin(x+y)-2\\[4pt]
-\sin(x+y)-2 & -\sin(x+y)+2
\end{pmatrix},
$$
where $\frac{\partial^2 f}{\partial x\partial y}=\frac{\partial^2 f}{\partial y\partial x}$.


```{r 1_2, echo=FALSE}
#1.2
## Gradient
grad_f <- function(x, y) {
  c(
    dx = cos(x + y) + 2*(x - y) - 1.5,
    dy = cos(x + y) - 2*(x - y) + 2.5
  )
}

## Hessian
hess_f <- function(x, y) {
  s <- -sin(x + y)
  matrix(
    c(s + 2, s - 2,
      s - 2, s + 2),
    nrow = 2, byrow = TRUE,
    dimnames = list(c("x", "y"), c("x", "y"))
  )
}

## wrappers that take par = c(x, y) for optimization
fn_par   <- function(par) f(par[1], par[2])
gr_par   <- function(par) grad_f(par[1], par[2])
hess_par <- function(par) hess_f(par[1], par[2])

## test
par0 <- c(0, 0)
fn_par(par0)
gr_par(par0)
hess_par(par0)
```

### 1.3 Newton algorithm

We are asked to search for the global minimum solution within the box 

$$
x \in [-1.5,\,4], \qquad y \in [-3,\,4].
$$

so the algorithm code needs to be adjusted to prevent the search path wandering outside this box. 

```{r 1_3, echo=FALSE}
#1.3

newton2d_box <- function(fn, gr, hess, x0, y0,
                         lower, upper,
                         tol = 1e-8, maxit = 100,
                         lambda0 = 1e-6, lambda_mult = 10,
                         alpha0 = 1, beta = 0.5, c1 = 1e-4,
                         ..., verbose = FALSE) {

  project <- function(par) pmin(pmax(par, lower), upper)

  ## infinity-norm of projected gradient (KKT-style for box constraints)
  projgrad_inf <- function(x, g, eps = 1e-12) {
    pg <- g
    for (i in 1:2) {
      if (x[i] <= lower[i] + eps) pg[i] <- min(g[i], 0)  # should be >=0 at lower bound
      if (x[i] >= upper[i] - eps) pg[i] <- max(g[i], 0)  # should be <=0 at upper bound 
    }
    max(abs(pg))
  }

  x <- project(c(x0, y0))
  f0 <- fn(x[1], x[2], ...)

  for (k in 1:maxit) {
    g <- gr(x[1], x[2], ...)
    pg_norm <- projgrad_inf(x, g)

    if (pg_norm < tol) {
      return(list(par = x, value = f0, iter = k - 1, converged = TRUE,
                  projgrad = pg_norm))
    }

    H <- hess(x[1], x[2], ...)
    I2 <- diag(2)

    ## Regularize Hessian until solvable
    lambda <- lambda0
    p <- NULL
    repeat {
      p_try <- tryCatch(solve(H + lambda * I2, g), error = function(e) NULL)
      if (!is.null(p_try)) { p <- p_try; break }
      lambda <- lambda * lambda_mult
      if (lambda > 1e12) stop("Could not compute Newton step (even after regularization).")
    }

    ## Backtracking line search with Armijo sufficient decrease
    a <- alpha0
    repeat {
      x_new <- project(x - a * p)
      f_new <- fn(x_new[1], x_new[2], ...)

      ## directional derivative along -p is approx -g^T p
      armijo_rhs <- f0 - c1 * a * sum(g * p)

      if (is.finite(f_new) && f_new <= armijo_rhs) break
      a <- a * beta
      if (a < 1e-12) break
    }

    if (verbose) {
      cat(sprintf(
        "iter=%d  x=%.8f y=%.8f  f=%.10f  step=%.1e  ||pg||_inf=%.3e  lambda=%.1e\n",
        k, x_new[1], x_new[2], f_new, a, pg_norm, lambda
      ))
    }

    ## Only use step-size convergence as a secondary criterion
    if (sqrt(sum((x_new - x)^2)) < tol && pg_norm < tol) {
      return(list(par = x_new, value = f_new, iter = k, converged = TRUE,
                  projgrad = pg_norm))
    }

    x <- x_new
    f0 <- f_new
  }

  list(par = x, value = f0, iter = maxit, converged = FALSE,
       projgrad = projgrad_inf(x, gr(x[1], x[2], ...)))
}


# newton2d_box <- function(fn, gr, hess, x0, y0,
#                          lower, upper,
#                          tol = 1e-8, maxit = 100,
#                          lambda0 = 1e-6, lambda_mult = 10,
#                          alpha0 = 1, beta = 0.5,
#                          ..., verbose = FALSE) {
# 
#   #To keep the results within the prescribed x,y limits
#   project <- function(par) pmin(pmax(par, lower), upper)
# 
#   x <- project(c(x0, y0))
#   f0 <- fn(x[1], x[2], ...)
# 
#   for (k in 1:maxit) {
#     g <- gr(x[1], x[2], ...)
#     H <- hess(x[1], x[2], ...)
#     I2 <- diag(2)
# 
#     ## regularize until solvable
#     lambda <- lambda0
#     p <- NULL
#     repeat {
#       p_try <- tryCatch(solve(H + lambda * I2, g), error = function(e) NULL)
#       if (!is.null(p_try)) { p <- p_try; break }
#       lambda <- lambda * lambda_mult
#       if (lambda > 1e12) stop("Could not compute Newton step.")
#     }
# 
#     ## backtracking line search with projection
#     a <- alpha0
#     repeat {
#       x_new <- project(x - a * p)
#       f_new <- fn(x_new[1], x_new[2], ...)
#       if (is.finite(f_new) && f_new <= f0) break
#       a <- a * beta
#       if (a < 1e-12) break
#     }
# 
#     if (verbose) {
#       cat(sprintf("iter=%d  x=%.8f y=%.8f  f=%.10f  step=%.1e\n",
#                   k, x_new[1], x_new[2], f_new, a))
#     }
# 
#     if (sqrt(sum((x_new - x)^2)) < tol) {
#       return(list(par = x_new, value = f_new, iter = k, converged = TRUE))
#     }
# 
#     x <- x_new
#     f0 <- f_new
#   }
# 
#   list(par = x, value = f0, iter = maxit, converged = FALSE)
# }

# ## x_{k+1} = x_k - H(x_k)^{-1} * g(x_k)
# newton2d_reg <- function(fn, gr, hess, x0, y0,
#                          tol = 1e-8, maxit = 100,
#                          lambda0 = 1e-6, lambda_mult = 10,
#                          ..., verbose = FALSE) {
# 
#   x <- c(x0, y0)
# 
#   for (k in 1:maxit) {
#     g <- gr(x[1], x[2], ...)
#     H <- hess(x[1], x[2], ...)
#     I2 <- diag(2)
# 
#     lambda <- lambda0
#     p <- NULL
# 
#     ## Keep increasing lambda until the linear system is solvable
#     repeat {
#       Hp <- H + lambda * I2
#       p_try <- tryCatch(solve(Hp, g), error = function(e) NULL)  # catch singularity
#       if (!is.null(p_try)) { p <- p_try; break }
#       lambda <- lambda * lambda_mult
#       if (lambda > 1e12) stop("Hessian remained singular/ill-conditioned even after regularization.")
#     }
# 
#     x_new <- x - p
# 
#     if (verbose) {
#       cat(sprintf("iter=%d  x=%.10f y=%.10f  f=%.10f  ||grad||=%.3e  lambda=%.1e\n",
#                   k, x_new[1], x_new[2], fn(x_new[1], x_new[2], ...),
#                   sqrt(sum(g^2)), lambda))
#     }
# 
#     if (sqrt(sum((x_new - x)^2)) < tol) {
#       return(list(par = x_new, value = fn(x_new[1], x_new[2], ...),
#                   iter = k, converged = TRUE))
#     }
#     x <- x_new
#   }
# 
#   list(par = x, value = fn(x[1], x[2], ...), iter = maxit, converged = FALSE)
# }
# 
lower <- c(-1.5, -3)
upper <- c( 4.0,  4)
res <- newton2d_box(fn = f, gr = grad_f, hess = hess_f, 
                    x0 = 0, y0 = 0, lower=lower, upper=upper, verbose = TRUE)
res$par
res$value
```

### 1.4 Testing with different starting points

```{r 1_4, echo=FALSE}
#1.4
lower <- c(-1.5, -3)
upper <- c( 4.0,  4)

one_run <- function(x0, y0) {
  nres <- newton2d_box(f, grad_f, hess_f, x0, y0, lower, upper)

  ores <- optim(
    par = c(x0, y0),
    fn = fn_par, gr = gr_par,
    method = "L-BFGS-B",
    lower = lower, upper = upper
  )  # box constraints supported by L-BFGS-B

  data.frame(
    x0 = x0, y0 = y0,
    newton_iter = nres$iter,
    newton_x = nres$par[1], newton_y = nres$par[2],
    newton_value = nres$value,
    optim_iter = ores$counts[["function"]],
    optim_x = ores$par[1], optim_y = ores$par[2],
    optim_value = ores$value
  )
}


starts <- rbind(
  c(-1.0, -2.0),
  c( 3.5,  3.5),
  c( 0.0,  0.0),
  c( 4.0, -3.0),
  c(-1.5,  4.0)
)
colnames(starts) <- c("x0", "y0")
# 
# one_run <- function(x0, y0) {
#   ## Newton (regularized)
#   nres <- newton2d_reg(fn = f, gr = grad_f, hess = hess_f, x0 = x0, y0 = y0)
# 
#   ## optim() comparison, BFGS with analytic gradient
#   ores <- optim(par = c(x0, y0), fn = fn_par, gr = gr_par, method = "BFGS")
# 
#   data.frame(
#     x0 = x0, y0 = y0,
#     newton_iter = nres$iter,
#     newton_x = unname(nres$par[1]),
#     newton_y = unname(nres$par[2]),
#     newton_value = nres$value,
#     optim_iter = ores$counts[["function"]],
#     optim_x = unname(ores$par[1]),
#     optim_y = unname(ores$par[2]),
#     optim_value = ores$value,
#     stringsAsFactors = FALSE
#   )
# }
# 
res_table <- do.call(
  rbind,
  lapply(seq_len(nrow(starts)), function(i) one_run(starts[i, 1], starts[i, 2]))
)

res_table_fmt <- res_table
num_cols <- names(res_table_fmt)[sapply(res_table_fmt, is.numeric)]

res_table_fmt[num_cols] <- lapply(res_table_fmt[num_cols], \(v) formatC(v, format = "f", digits = 6))  # fixed decimals
print(res_table_fmt, row.names = FALSE)

## distinct solutions reached (rounded to group near-equal points)
unique(round(res_table[, c("newton_x", "newton_y")], 6))
unique(round(res_table[, c("optim_x", "optim_y")], 6))

```

## 2


## Appendix


```{r, ref.label=setdiff(knitr::all_labels(), c('setup')), echo=TRUE, eval=FALSE}



```
