---
title: "Lab 1"
author: "Aron Enström (aroen488), Sergey Volchkov (servo519)"
date: "2026-01-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1


### 1.1. Function plot
```{r 1_1, echo=FALSE}

#1.1
library(ggplot2)
g <- function(x) log(x + 1) / (x^(3/2) + 1)

df <- data.frame(x = seq(0, 4, length.out = 1000))
df$g <- g(df$x)

ggplot(df, aes(x, g)) +
  geom_line(linewidth = 1) +
  labs(
    title = "g(x) = log(x+1)/(x^(3/2)+1)",
    x = "x", y = "g(x)"
  ) +
  theme_minimal()
```

The function maximum should be at around 1.

### 1.2. Derivative and plot

$g(x)=\dfrac{\log(x+1)}{x^{3/2}+1}$.

We set $u(x)=\log(x+1)$ and $v(x)=x^{3/2}+1$, so $g(x)=\dfrac{u(x)}{v(x)}$.

By the quotient rule,
$$
g'(x)=\frac{u'(x)v(x)-u(x)v'(x)}{(v(x))^2}.
$$

Then $u'(x)=\dfrac{1}{x+1}$ and $v'(x)=\dfrac{3}{2}x^{1/2}$.

Which gives:
$$
g'(x)=\frac{\frac{1}{x+1}(x^{3/2}+1)-\log(x+1)\left(\frac{3}{2}x^{1/2}\right)}{(x^{3/2}+1)^2}.
$$ 

Or equivalently,
$$
g'(x)=\frac{(x^{3/2}+1)-\frac{3}{2}x^{1/2}(x+1)\log(x+1)}{(x+1)(x^{3/2}+1)^2}.
$$ 

```{r 1_2, echo=FALSE}
#1.2
g_prime <- function(x) {
  ((x^(3/2) + 1)/(x + 1) - (3/2) * x^(1/2) * log(x + 1)) / (x^(3/2) + 1)^2
}

curve(g_prime(x), from = 0, to = 4, n = 1000,
      xlab = "x", ylab = "g'(x)",
      main = "Derivative g'(x) on [0, 4]")

abline(h = 0, lty = 2, col = "red")
grid()
```

### 1.3. Bisection method implementation

```{r 1_3, echo=FALSE}
#1.3

## Bisection method (root-finding) for interval [a, b]
bisect_root <- function(f, a, b, tol = 1e-10, max_iter = 200) {
  fa <- f(a); fb <- f(b)

  if (!is.finite(fa) || !is.finite(fb))
    stop("f(a) or f(b) is not finite.")

  if (fa == 0) return(list(root = a, f_root = fa, iter = 0, a = a, b = b))
  if (fb == 0) return(list(root = b, f_root = fb, iter = 0, a = a, b = b))

  ## Need a sign change for bisection
  if (fa * fb > 0)
    stop("No sign change on [a, b]. Choose an interval where f(a) and f(b) have opposite signs.")

  for (iter in 1:max_iter) {
    c <- (a + b) / 2
    fc <- f(c)

    if (!is.finite(fc))
      stop("f(midpoint) not finite; try a different interval.")

    if (abs(fc) < tol || (b - a) / 2 < tol)
      return(list(root = c, f_root = fc, iter = iter, a = a, b = b))

    if (fa * fc < 0) {
      b <- c; fb <- fc
    } else {
      a <- c; fa <- fc
    }
  }

  list(root = (a + b)/2, f_root = f((a + b)/2), iter = max_iter, a = a, b = b)
}

## Find a critical point of g by solving g'(x)=0 on a user-selected interval
find_local_max_g <- function(a, b, tol = 1e-10, max_iter = 200) {
  out <- bisect_root(g_prime, a, b, tol = tol, max_iter = max_iter)
  x_star <- out$root

  ## First-derivative sign change test: + to - indicates local max
  eps <- min(1e-4, (b - a) / 100)
  left  <- g_prime(max(a, x_star - eps))
  right <- g_prime(min(b, x_star + eps))

  type <- if (is.finite(left) && is.finite(right) && left > 0 && right < 0) {
    "local maximum (g' changes + to -)"
  } else if (is.finite(left) && is.finite(right) && left < 0 && right > 0) {
    "local minimum (g' changes - to +)"
  } else {
    "critical point found; sign-change unclear on chosen interval"
  }

  list(x_star = x_star, g_at_x_star = g(x_star), type = type, details = out)
}

## Test
a = 0 
b = 4
res <- find_local_max_g(a = a, b = b)
```

Test results for $g(x)$ using the bisection method on the interval [`r a`, `r b`] : 

```{r 1_3_1, echo=FALSE}
res
```

### 1.4. Secant method implementation

```{r 1_4, echo=FALSE}
#1.4
## Secant method for root finding
secant_root <- function(f, x0, x1, tol = 1e-10, max_iter = 200) {
  f0 <- f(x0); f1 <- f(x1)

  if (!is.finite(f0) || !is.finite(f1))
    stop("f(x0) or f(x1) is not finite.")

  if (f0 == 0) return(list(root = x0, f_root = f0, iter = 0, x0 = x0, x1 = x1))
  if (f1 == 0) return(list(root = x1, f_root = f1, iter = 0, x0 = x0, x1 = x1))

  for (iter in 1:max_iter) {
    denom <- (f1 - f0)
    if (denom == 0) stop("Secant breakdown: f(x1) - f(x0) = 0.")

    ## x_{n+1} = x_n - f(x_n) (x_n - x_{n-1}) / (f(x_n) - f(x_{n-1}))
    x2 <- x1 - f1 * (x1 - x0) / denom

    if (!is.finite(x2)) stop("Non-finite iterate produced; try different starting values.")

    if (abs(x2 - x1) < tol)
      return(list(root = x2, f_root = f(x2), iter = iter, x0 = x0, x1 = x1))

    ## shift (x0,f0) <- (x1,f1), (x1,f1) <- (x2,f2)
    x0 <- x1; f0 <- f1
    x1 <- x2; f1 <- f(x1)

    if (!is.finite(f1)) stop("Non-finite f(x) encountered; try different starting values.")
    if (abs(f1) < tol)  # optional early stop on function value
      return(list(root = x1, f_root = f1, iter = iter, x0 = x0, x1 = x1))
  }

  list(root = x1, f_root = f1, iter = max_iter, x0 = x0, x1 = x1)
}

## Find local max of g by applying secant to g'
find_local_max_g_secant <- function(x0, x1, tol = 1e-10, max_iter = 200) {
  out <- secant_root(g_prime, x0, x1, tol = tol, max_iter = max_iter)
  x_star <- out$root

  ## First-derivative sign-change check (heuristic)
  eps <- min(1e-4, abs(x1 - x0) / 100)
  left  <- g_prime(x_star - eps)
  right <- g_prime(x_star + eps)

  type <- if (is.finite(left) && is.finite(right) && left > 0 && right < 0) {
    "local maximum (g' changes + to -)"
  } else if (is.finite(left) && is.finite(right) && left < 0 && right > 0) {
    "local minimum (g' changes - to +)"
  } else {
    "critical point found; sign-change unclear from starting values"
  }

  list(x_star = x_star, g_at_x_star = g(x_star), type = type, details = out)
}

## Test
x0 = 0.2
x1 = 1
res <- find_local_max_g_secant(x0 = x0, x1 = x1)
```

Test results for $g(x)$ using the secant method on the interval [`r x0`, `r x1`] : 

```{r 1_4_1, echo=FALSE}
res
```
### 1.5. Comparison of the two methods for the task at hand

```{r 1_5, echo=FALSE, warning=FALSE}
## 1.5
# The "true" maximizer x* on [0,4]
# Find sign changes of g' on a fine grid, then bisect in the first bracket.
grid <- seq(0, 4, length.out = 2001)
vals <- g_prime(grid)

idx <- which(diff(sign(vals)) != 0)  # sign changes
if (length(idx) == 0) stop("No sign change found on [0,4]; cannot bracket a root.")

a_star <- grid[idx[1]]
b_star <- grid[idx[1] + 1]
true_out <- bisect_root(g_prime, a_star, b_star, tol = 1e-14, max_iter = 2000)
x_true <- true_out$root

## Bisection: test many [a,b]
bisect_grid <- expand.grid(
  a = seq(0, 3.5, by = 0.5),
  b = seq(0.5, 4.0, by = 0.5)
) 
bisect_grid <- subset(bisect_grid, a < b)

## Secant: test many (x0,x1) pairs
secant_grid <- expand.grid(
  x0 = seq(0.0, 3.5, by = 0.5),
  x1 = seq(0.5, 4.0, by = 0.5)
)  
secant_grid <- subset(secant_grid, x0 < x1)

## Helper: run safely and return a row
run_one <- function(method, p1, p2, tol_root = 1e-4) {
  if (method == "bisection") {
    out <- tryCatch(bisect_root(g_prime, p1, p2), error = function(e) NULL)
    root <- if (is.null(out)) NA_real_ else out$root
    iter <- if (is.null(out)) NA_integer_ else out$iter
    status <- if (is.null(out)) "fail" else "ok"
    data.frame(method = method, start1 = p1, start2 = p2,
               status = status,
               root = root,
               true_root_found = isTRUE(!is.na(root) && abs(root - x_true) <= tol_root),
               iterations = iter)
  } else {
    out <- tryCatch(secant_root(g_prime, p1, p2), error = function(e) NULL)
    root <- if (is.null(out)) NA_real_ else out$root
    iter <- if (is.null(out)) NA_integer_ else out$iter
    status <- if (is.null(out)) "fail" else "ok"
    data.frame(method = method, start1 = p1, start2 = p2,
               status = status,
               root = root,
               true_root_found = isTRUE(!is.na(root) && abs(root - x_true) <= tol_root),
               iterations = iter)
  }
}

## Run and bind results
bisect_res <- do.call(rbind, lapply(seq_len(nrow(bisect_grid)), function(i)
  run_one("bisection", bisect_grid$a[i], bisect_grid$b[i])
)) 

secant_res <- do.call(rbind, lapply(seq_len(nrow(secant_grid)), function(i)
  run_one("secant", secant_grid$x0[i], secant_grid$x1[i])
))

results <- rbind(bisect_res, secant_res)

## Sort for readability
results <- results[order(results$method, results$start1, results$start2), ]

## Print summary table
results
```
The bisection method fails when the starting interval does not include the true root.

The secant method can succeed even when the starting interval does not include the true root and takes fewer iterations (up to 12 vs around 30 for the bisection method). 

Using the update formula for the secant method

$$
x_{n+1}
= x_n - f(x_n)\,\frac{x_n - x_{n-1}}{f(x_n)-f(x_{n-1})}.
$$

which in our case is equivalent to

$$
x_{n+1}
= x_n - g'(x_n)\,\frac{x_n - x_{n-1}}{g'(x_n)-g'(x_{n-1})}.
$$

The pattern in the table for the secant method (lots of roots around $10^4$ or “fail”) is consistent with secant taking a bad step and then wandering to a region where $g'(x)$ is very small/flat (we can observe it in the region of the $g'(x)$ graph in section 1.2 for in the $x$ interval from 1.5 to 4 and we can see from the table that the algorithm fails in nearly all cases when one or both starting values are in this interval) or where the update becomes ill-conditioned (e.g. $g'(x_n)-g'(x_{n-1})$ tiny), which can blow up the update and lead to divergence or stagnation. It appears that the secant method can diverge when it encounters a near-zero slope / derivative-like quantity in the update.

The amount of programming effort is perhaps slightly higher for the secant method.


### 1.6. General comparison of the bisection and secant methods

For the task at hand, we would choose the bisection method despite it being slower than the secant method, we can determine the region containing the true root quite reliably, while the secant method appears to be rather sensitive to the choice of the starting values due to the flatness of the derivative function. 

Generally, we prefer bisection when we have a bracketing interval $[a,b]$ with a sign change 
$f(a)f(b)<0$, and we want a method that is guaranteed to converge under mild conditions (continuous $f$). We prefer secant when we have good starting values (close to the root), function evaluations are costly and we want fewer iterations than bisection.  

## 2

### 2.1. Function to calculate determinant of the information matrix

```{r 2_1, echo=FALSE}
# 2.1
f <- function(a) {
  x <- c(-1, -a, a, 1)

  X <- cbind(
    1,
    x,
    x^2,
    x^3
  )

  det(t(X) %*% X)
}
```

### 2.2. Function plot

```{r 2_2, echo=FALSE}
# 2.2
f <- function(a) {
  x <- c(-1, -a, a, 1)
  X <- cbind(1, x, x^2, x^3)
  det(t(X) %*% X)
}

aa <- seq(0, 1, length.out = 2001)
ff <- sapply(aa, f)

plot(aa, ff, type = "l", lwd = 2,
     xlab = "a", ylab = "f(a) = det(t(X) %*% X)",
     main = "Determinant of information matrix vs a")

abline(v = c(0, 1), lty = 2, col = "grey50")
points(c(0, 1), c(f(0), f(1)), pch = 19, col = "red")

f0 <- f(0); f1 <- f(1)
```

#### Interpretation of a=0 and a=1

- Case $a=0$: the design points are $x=(-1,0,0,1)$, so $x=0$ is duplicated and there are only three distinct $x$-values. This typically makes the design matrix $X$ lose rank, so the information matrix $X^\top X$ becomes singular and $f(0)=\det(X^\top X)=0$, meaning the cubic regression parameters are not all identifiable from this design.

- Case $a=1$: the design points are $x=(-1,-1,1,1)$, so there are only two distinct $x$-values (each duplicated). Then $X$ cannot have full column rank $4$, so $X^\top X$ is singular and $f(1)=\det(X^\top X)=0$.


### 2.3. Local maximum of the function

```{r 2_3, echo=FALSE}
# 2.3
opt <- optimize(f, interval = c(0, 1), maximum = TRUE)
```

The function has its maximum at `r opt$maximum` with the maximum value of `r opt$objective`. This matches what we see on the plot in 2.2. 


## Appendix


```{r, ref.label=setdiff(knitr::all_labels(), c('setup')), echo=TRUE, eval=FALSE}



```
