---
title: "Lab 4"
author: "Aron Enström (aroen488), Sergey Volchkov (servo519)"
date: "2026-02-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1: 


### a. Metropolis-Hastings with normal distribution as proposal

```{r 1_1, echo=FALSE, warning=FALSE}
#####
# 1(a)
#####

## Target: f(x) = (1/(2.4*10^6)) * x^4 * exp(-x/10), x > 0
## This is exactly Gamma(shape=5, scale=10).
shape_true <- 5
scale_true <- 10

## Use log of *unnormalized* density (constant cancels in MH ratio)
log_f_unnorm <- function(x) {
  ifelse(x > 0, 4 * log(x) - x / 10, -Inf)
}

## Generic MH sampler for arbitrary proposal q(y|x)
mh_sampler <- function(n = 10000, x0 = 1, rprop, dprop, ...) {
  x <- numeric(n)
  x[1] <- x0
  acc <- 0L

  for (t in 2:n) {
    xt <- x[t - 1]
    y  <- rprop(xt, ...)

    log_num <- log_f_unnorm(y)  + log(dprop(xt, y, ...))  # f(y) q(x|y)
    log_den <- log_f_unnorm(xt) + log(dprop(y, xt, ...))  # f(x) q(y|x)
    log_alpha <- log_num - log_den

    if (log(runif(1)) < log_alpha) {
      x[t] <- y
      acc <- acc + 1L
    } else {
      x[t] <- xt
    }
  }

  list(x = x, acc_rate = acc / (n - 1))
}

## Proposal "factories"
make_norm_rw_proposal <- function(sd = 1) {
  list(
    name = sprintf("Normal RW (sd=%g)", sd),
    rprop = function(x) rnorm(1, mean = x, sd = sd),
    dprop = function(y, x) dnorm(y, mean = x, sd = sd)
  )
}

make_chisq_floor_proposal <- function() {
  list(
    name = "Chi-square df = floor(x)+1",
    rprop = function(x) {
      df <- max(1L, floor(x) + 1L)     # floor(x+1) == floor(x)+1
      rchisq(1, df = df)
    },
    dprop = function(y, x) {
      df <- max(1L, floor(x) + 1L)
      dchisq(y, df = df)
    }
  )
}

run_sampler <- function(proposal, burn = 0, n = 10000, x0 = 1) {
  res <- mh_sampler(n = n, x0 = x0, rprop = proposal$rprop, dprop = proposal$dprop)
  x <- res$x

  cat("Proposal:", proposal$name, "\n")
  cat("Acceptance rate:", res$acc_rate, "\n")

  ## Trace + running mean
  par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))

  plot(x, type = "l", col = "gray30", xlab = "Iteration", ylab = "X_t",
       main = paste("Trace plot (", proposal$name, ")", sep = ""))

  run_mean <- cumsum(x) / seq_along(x)
  theo_mean <- shape_true * scale_true
  yl <- range(c(run_mean, theo_mean), finite = TRUE)

  plot(run_mean, type = "l", col = "steelblue4",
       xlab = "Iteration", ylab = "Running mean",
       main = "Running mean diagnostic", ylim = yl)
  abline(h = theo_mean, col = "red", lwd = 2, lty = 2)
  abline(v = burn,     col = "darkorange3", lwd = 2, lty = 2)

  legend("topright",
         legend = c("Running mean", "Theoretical mean", "Burn-in cutoff"),
         col    = c("steelblue4", "red", "darkorange3"),
         lty    = c(1, 2, 2),
         lwd    = c(1, 2, 2),
         bty    = "n")

  ## Post burn-in sample
  x_post <- if (burn > 0) x[-seq_len(burn)] else x

  ## Histogram + true density overlay
  par(mfrow = c(1, 1), mar = c(4, 4, 2, 1))
  hist(x_post, breaks = 50, freq = FALSE, col = "gray85", border = "white",
       main = sprintf("Samples after burn-in (burn = %d)", burn),
       xlab = "x")

  curve(dgamma(x, shape = shape_true, scale = scale_true),
        from = 0, to = max(x_post),
        add = TRUE, col = "red", lwd = 2)

  abline(v = theo_mean, col = "red", lwd = 2, lty = 2)
  
  x_post <- if (burn > 0) x[-seq_len(burn)] else x

  invisible(list(
    x = x,
    x_post = x_post,
    acc_rate = res$acc_rate,
    burn = burn,
    proposal = proposal$name
  ))
}
set.seed(1)
prop_norm <- make_norm_rw_proposal(sd = 1)
out_norm_sd1 <- run_sampler(prop_norm, burn = 0, n = 10000, x0 = 1)
x_norm_sd1 <- out_norm_sd1$x_post
```
With the standard deviation of 1 and not discarding the samples obtained during the burn-in, we are not seeing convergence to the true mean. The acceptance rate is too high (near 1), which indicates we are moving too slowly. 

### b. Metropolis-Hastings with chi-square distribution as proposal

```{r 1_2, echo=FALSE, warning=FALSE}
#####
# 1(b)
#####
set.seed(1)
prop_chi <- make_chisq_floor_proposal()
out_chi <- run_sampler(prop_chi, burn = 0, n = 10000, x0 = 1)
x_chi <- out_chi$x_post
```

### c. Metropolis-Hastings with normal distribution as proposal (sd=8)

```{r 1_3, echo=FALSE, warning=FALSE}
#####
# 1(c)
#####
set.seed(1)
prop_norm <- make_norm_rw_proposal(sd = 8)
out_norm_sd8 <- run_sampler(prop_norm, burn = 2000, n = 12000, x0 = 1)
x_norm_sd8 <- out_norm_sd8$x_post
```

### d. Comparison of the three methods

For `sd = 8` and the burn-in period of 2000, we are getting much better picture than with the originally used sd of 1. Our resulting histogram matches the gamma distribution much more closely and the acceptance rate is much lower. However, the chi-squared proposal still seems to be the best in terms of the speed of convergence.The chi-square proposal is state-dependent, e.g. $Y \mid X_t=x \sim \chi^2_{\lfloor x \rfloor + 1}$, so its typical proposal size changes with the current location of the chain. This often yields larger, more effective moves when the chain is in higher-density regions away from 0, so the chain can reach and traverse the typical set in fewer iterations than a fixed-scale random-walk. In contrast, a Normal random-walk proposal $Y \mid X_t=x \sim N(x,\sigma^2)$ has a constant step scale $\sigma$, which can lead to very slow exploration when $\sigma$ is too small (even if acceptance is very high). 

### e. Comparison of E(X) and medians obtained using the three methods

```{r 1_5, echo=FALSE, warning=FALSE}
#####
# 1(e)
#####
summ_tab <- data.frame(
  Chain  = c("Normal RW (sd=1)", "Normal RW (sd=8)", "Chi-square df=floor(x)+1"),
  Mean   = c(mean(x_norm_sd1, na.rm = TRUE),
             mean(x_norm_sd8, na.rm = TRUE),
             mean(x_chi,      na.rm = TRUE)),
  Median = c(median(x_norm_sd1, na.rm = TRUE),
             median(x_norm_sd8, na.rm = TRUE),
             median(x_chi,      na.rm = TRUE))
)

knitr::kable(summ_tab, digits = 3, caption = "Mean and median of MCMC samples")
```
### f. Theoretical E(X)

For a Gamma distribution parameterized as $X \sim \mathrm{Gamma}(\alpha,\theta)$ (shape $\alpha$, scale $\theta$), the expectation is $E[X]=\alpha\theta$. Thus for $X \sim \mathrm{Gamma}(5,10)$, $E[X]=5\cdot 10=50$ which is the value we used on our plots above. We can see that the chi-square method gets quite close to it (so does normal with sd = 8 but only after discarding the first 2000 samples).

We can also double-check the median in R:

```{r 1_6, echo=FALSE, warning=FALSE}
#####
# 1.6
#####
qgamma(0.5, shape=5, scale=10)
```

## Q2: 


### a. Ellipse plot

```{r 2_1, echo=FALSE, warning=FALSE}
#####
# 2(a)
#####
w  <- 1.999

a <- 1 - w^2/4
xmax <- 1/sqrt(a)
xv <- seq(-xmax, xmax, length.out = 4000)

disc <- pmax(0, 1 - a * xv^2)  # guard against tiny negatives from rounding
ylo  <- -(w/2) * xv - sqrt(disc)
yhi  <- -(w/2) * xv + sqrt(disc)

plot(xv, ylo, type = "n",
     xlab = expression(x[1]), ylab = expression(x[2]), las = 1,
     asp = 1,
     xlim = range(xv), ylim = range(c(ylo, yhi)))

lines(xv, ylo, lwd = 2, col = 8)
lines(xv, yhi, lwd = 2, col = 8)

```


### b. Conditional distributions

We have a bivariate random vector $X=(X_1,X_2)$ that is uniform on the region
$$
R_w=\{(x_1,x_2)\in\mathbb{R}^2:\; x_1^2+w x_1 x_2+x_2^2<1\},\qquad |w|<2.
$$

If $(X_1,X_2)$ is uniform on a 2D region $R$, then the conditional distribution of $X_1$ given $X_2=x_2$ is uniform on the horizontal slice $T_{x_2}=\{x_1:(x_1,x_2)\in R\}$ (and similarly with the roles swapped). 
Equivalently, the conditional density is constant on that slice and equals $1/\mathrm{length}(T_{x_2})$. 

#### Conditional of $X_1$ given $X_2=x_2$

Fix $x_2$. The constraint $x_1^2+w x_1 x_2+x_2^2<1$ is a quadratic inequality in $x_1$:
$$
x_1^2+(w x_2)\,x_1+(x_2^2-1)<0.
$$

Its discriminant is
$$
\Delta_1(x_2)=(w x_2)^2-4(x_2^2-1)=4-(4-w^2)x_2^2.
$$

So the slice is non-empty iff $\Delta_1(x_2)>0$, i.e.
$$
|x_2|<\frac{2}{\sqrt{4-w^2}}.
$$

When $\Delta_1(x_2)>0$, the admissible interval for $x_1$ is
$$
\ell_1(x_2)=\frac{-w x_2-\sqrt{\Delta_1(x_2)}}{2},
\qquad
u_1(x_2)=\frac{-w x_2+\sqrt{\Delta_1(x_2)}}{2}.
$$

Therefore,
$$
X_1\mid (X_2=x_2)\sim \mathrm{Unif}\big(\ell_1(x_2),u_1(x_2)\big), \quad \text{for } |x_2|<\frac{2}{\sqrt{4-w^2}}.
$$
because a uniform distribution on an interval has constant density on that interval.

The conditional density can be written as
$$
f_{X_1\mid X_2}(x_1\mid x_2)=
\begin{cases}
\frac{1}{u_1(x_2)-\ell_1(x_2)}=\frac{1}{\sqrt{\Delta_1(x_2)}}, & \ell_1(x_2)\le x_1\le u_1(x_2),\\
0, & \text{otherwise}.
\end{cases}
$$
using that a $\mathrm{Unif}(a,b)$ density is $1/(b-a)$ on $[a,b]$.

Similarly,

$$
X_2\mid (X_1=x_1)\sim \mathrm{Unif}\big(\ell_2(x_1),u_2(x_1)\big), \quad \text{for } |x_1|<\frac{2}{\sqrt{4-w^2}}.
$$
by the same “uniform on a slice” principle.


### c. Gibbs sampling

```{r 2_3, echo=FALSE, warning=FALSE}
#####
# 2(c)
#####
w <- 1.999
tol <- 1e-12

Q <- function(x1, x2, w) x1^2 + w * x1 * x2 + x2^2

bounds_x1_given_x2 <- function(x2, w, tol = 1e-12) {
  Delta <- 4 - (4 - w^2) * x2^2
  if (Delta < -tol) stop("Infeasible x2 (outside support): discriminant < 0.")
  Delta <- max(0, Delta)
  s <- sqrt(Delta)
  lo <- (-w * x2 - s) / 2
  hi <- (-w * x2 + s) / 2
  c(lower = min(lo, hi), upper = max(lo, hi))
}

bounds_x2_given_x1 <- function(x1, w, tol = 1e-12) {
  Delta <- 4 - (4 - w^2) * x1^2
  if (Delta < -tol) stop("Infeasible x1 (outside support): discriminant < 0.")
  Delta <- max(0, Delta)
  s <- sqrt(Delta)
  lo <- (-w * x1 - s) / 2
  hi <- (-w * x1 + s) / 2
  c(lower = min(lo, hi), upper = max(lo, hi))
}

gibbs_Rw <- function(n = 1000, burn = 2000, w = 1.999, x0 = c(0, 0), tol = 1e-12) {
  x1 <- x0[1]; x2 <- x0[2]
  if (Q(x1, x2, w) >= 1) stop("Initial state x0 must satisfy Q(x0) < 1.")

  out <- matrix(NA_real_, nrow = n, ncol = 2)
  colnames(out) <- c("X1", "X2")

  for (i in seq_len(burn + n)) {
    b1 <- bounds_x1_given_x2(x2, w, tol)
    x1 <- runif(1, min = b1["lower"], max = b1["upper"])

    b2 <- bounds_x2_given_x1(x1, w, tol)
    x2 <- runif(1, min = b2["lower"], max = b2["upper"])

    # Safety: if rounding nudges us outside, scale back inside (Q(c x)=c^2 Q(x))
    q <- Q(x1, x2, w)
    if (!is.finite(q)) stop("Non-finite Q encountered.")
    if (q >= 1) {
      cscale <- sqrt((1 - 10 * tol) / q)
      x1 <- x1 * cscale
      x2 <- x2 * cscale
    }

    if (i > burn) out[i - burn, ] <- c(x1, x2)
  }
  out
}

set.seed(1)
samp <- gibbs_Rw(n = 1000, burn = 5000, w = w)

a <- 1 - w^2 / 4
xmax <- 1 / sqrt(a)
xv <- seq(-xmax, xmax, length.out = 4000)
disc <- pmax(0, 1 - a * xv^2)
ylo  <- -(w / 2) * xv - sqrt(disc)
yhi  <- -(w / 2) * xv + sqrt(disc)

plot(xv, ylo, type = "n",
     xlab = expression(x[1]), ylab = expression(x[2]),
     asp = 1, las = 1,
     xlim = range(xv), ylim = range(c(ylo, yhi)))
lines(xv, ylo, lwd = 2, col = 8)
lines(xv, yhi, lwd = 2, col = 8)
points(samp[, "X1"], samp[, "X2"], pch = 16, cex = 0.5, col = rgb(0, 0, 0, 0.35))

# Estimate P(X1 > 0)
p_hat <- mean(samp[, "X1"] > 0)
cat("P(X1 > 0):", p_hat)

set.seed(2)
reps <- 5
p_hats <- replicate(reps, {
  s <- gibbs_Rw(n = 1000, burn = 5000, w = w)
  mean(s[, "X1"] > 0)
})
cat("P(X1 > 0) based on repeated sampling:", p_hats)

```
By symmetry, the true value of $P(X1 > 0)$ should be 0.5.

### d. Why the results are worse than for w=1.8

Gibbs sampling is less successful at $w=1.999$ because the target region becomes extremely elongated (nearly “degenerate”), so $X_1$ and $X_2$ are very strongly coupled and single-coordinate updates move only tiny distances along the long, thin direction—leading to very slow mixing and high autocorrelation.

### e. Results for transformed distribution U = (U1,U2) = (X1 - X2,X1 + X2)

```{r 2_5, echo=FALSE, warning=FALSE}
#####
# 2(e)
#####
w    <- 1.999
n    <- 1000
burn <- 5000
tol  <- 1e-12

Qx <- function(x1, x2, w) x1^2 + w * x1 * x2 + x2^2
Qu <- function(u1, u2, w) (2 - w) * u1^2 + (2 + w) * u2^2

X_from_U <- function(u1, u2) c(x1 = (u2 + u1) / 2, x2 = (u2 - u1) / 2)
U_from_X <- function(x1, x2) c(u1 = x1 - x2, u2 = x1 + x2)

a1 <- sqrt(4 / (2 - w))  # max |u1|
a2 <- sqrt(4 / (2 + w))  # max |u2|

tt <- seq(0, 2 * pi, length.out = 4000)
u1_boundary <- a1 * cos(tt)
u2_boundary <- a2 * sin(tt)

plot(u1_boundary, u2_boundary, type = "l", lwd = 2, col = 8,
     xlab = expression(U[1]), ylab = expression(U[2]),
     asp = 1, las = 1)

bounds_u1_given_u2 <- function(u2, w, tol = 1e-12) {
  denom <- (2 - w)
  num   <- 4 - (2 + w) * u2^2
  if (num < -tol) stop("Infeasible u2 (outside support).")
  num <- max(0, num)
  r <- sqrt(num / denom)
  c(lower = -r, upper = r)
}

bounds_u2_given_u1 <- function(u1, w, tol = 1e-12) {
  denom <- (2 + w)
  num   <- 4 - (2 - w) * u1^2
  if (num < -tol) stop("Infeasible u1 (outside support).")
  num <- max(0, num)
  r <- sqrt(num / denom)
  c(lower = -r, upper = r)
}

gibbs_U <- function(n = 1000, burn = 5000, w = 1.999, u0 = c(0, 0), tol = 1e-12) {
  u1 <- u0[1]; u2 <- u0[2]
  if (Qu(u1, u2, w) >= 4) stop("Initial u0 must satisfy (2-w)u1^2 + (2+w)u2^2 < 4.")

  out <- matrix(NA_real_, nrow = n, ncol = 2)
  colnames(out) <- c("U1", "U2")

  for (i in seq_len(burn + n)) {
    b1 <- bounds_u1_given_u2(u2, w, tol)
    u1 <- runif(1, min = b1["lower"], max = b1["upper"])

    b2 <- bounds_u2_given_u1(u1, w, tol)
    u2 <- runif(1, min = b2["lower"], max = b2["upper"])

    # Numerical safety: nudge strictly inside if needed (scale (u1,u2) since Qu is quadratic)
    q <- Qu(u1, u2, w)
    if (!is.finite(q)) stop("Non-finite quadratic form encountered.")
    if (q >= 4) {
      cscale <- sqrt((4 - 10 * tol) / q)
      u1 <- u1 * cscale
      u2 <- u2 * cscale
    }

    if (i > burn) out[i - burn, ] <- c(u1, u2)
  }
  out
}

set.seed(1)
Usamp <- gibbs_U(n = n, burn = burn, w = w)

points(Usamp[, "U1"], Usamp[, "U2"], pch = 16, cex = 0.5, col = rgb(0, 0, 0, 0.35))

## Estimate P(X1 > 0) using U
# X1 = (U2 + U1)/2, so event X1 > 0 is equivalent to U1 + U2 > 0.
p_hat_U <- mean((Usamp[, "U1"] + Usamp[, "U2"]) > 0)
cat("P(X1 > 0) as P((U2 + U1)/2 > 0):", p_hat_U)

set.seed(123)

reps <- 10
p_hats_U <- replicate(reps, {
  U <- gibbs_U(n = 1000, burn = 5000, w = w)
  mean((U[, "U1"] + U[, "U2"]) > 0)
})

cat("P(X1 > 0) as P((U2 + U1)/2 > 0) from multiple runs:", p_hats_U)
```
We can see that the results are much closer to the true value of $P(X1 > 0)$ which is equal to 0.5. The estimates are closer to 0.5 because the transformation to $U_1=X_1-X_2,U_2=X_1+X_2$ rotates the geometry by 45 degrees so the support becomes an axis-aligned ellipse, and then coordinate-wise Gibbs updates are aligned with the ellipse’s principal axes so the chain moves efficiently.


## Appendix


```{r, ref.label=setdiff(knitr::all_labels(), c('setup')), echo=TRUE, eval=FALSE}



```
